%  ========================================================================
%  Copyright (c) 1985 The University of Washington
%
%  Licensed under the Apache License, Version 2.0 (the "License");
%  you may not use this file except in compliance with the License.
%  You may obtain a copy of the License at
%
%      http://www.apache.org/licenses/LICENSE-2.0
%
%  Unless required by applicable law or agreed to in writing, software
%  distributed under the License is distributed on an "AS IS" BASIS,
%  WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
%  See the License for the specific language governing permissions and
%  limitations under the License.
%  ========================================================================
%

% Documentation for University of Washington thesis LaTeX document class
% by Jim Fox
% fox@washington.edu
%
%    Revised 2020/02/24, added \caption()[]{} option.  No ToC.
%
%    Revised for version 2015/03/03 of uwthesis.cls
%    Revised, 2016/11/22, for cleanup of sample copyright and title pages
%
%    This document is contained in a single file ONLY because
%    I wanted to be able to distribute it easily.  A real thesis ought
%    to be contained on many files (e.g., one for each chapter, at least).
%
%    To help you identify the files and sections in this large file
%    I use the string '==========' to identify new files.
%
%    To help you ignore the unusual things I do with this sample document
%    I try to use the notation
%       
%    % --- sample stuff only -----
%    special stuff for my document, but you don't need it in your thesis
%    % --- end-of-sample-stuff ---


%    Printed in twoside style now that that's allowed
%
 
\documentclass [11pt, proquest] {uwthesis}[2020/02/24]

\usepackage{graphicx}
\usepackage{tabularx}
 
%
% The following line would print the thesis in a postscript font 

% \usepackage{natbib}
% \def\bibpreamble{\protect\addcontentsline{toc}{chapter}{Bibliography}}

\setcounter{tocdepth}{2}  % Print the chapter and sections to the toc
 

% ==========   Local defs and mods
%

% --- sample stuff only -----
% These format the sample code in this document

\usepackage{alltt}  % 
\newenvironment{demo}
  {\begin{alltt}\leftskip3em
     \def\\{\ttfamily\char`\\}%
     \def\{{\ttfamily\char`\{}%
     \def\}{\ttfamily\char`\}}}
  {\end{alltt}}
 
% metafont font.  If logo not available, use the second form
%
% \font\mffont=logosl10 scaled\magstep1
\let\mffont=\sf
% --- end-of-sample-stuff ---
 



\begin{document}
 
% ==========   Preliminary pages
%
% ( revised 2012 for electronic submission )
%

\prelimpages
 
%
% ----- copyright and title pages
%
\Title{Planetesimal Accretion in the Solar System and Beyond: An Application to Systems of Tightly-Packed Inner Planets}
\Author{Spencer Wallace}
\Year{2023}
\Program{Astronomy}

\Chair{Thomas Quinn}{}{Astronomy}
\Signature{Rory Barnes}
\Signature{Eric Agol}

\copyrightpage

\titlepage  

 
%
% ----- signature and quoteslip are gone
%

%
% ----- abstract
%


\setcounter{page}{-1}
\abstract{%
Planetesimals are the smallest gravitationally bound objects to play a role in the planet formation process. In a bottom-up fashion, these bodies are thought to collide and grow to form protoplanets, which are roughly Mars-sized objects, and then eventually coalesce into larger worlds. Throughout this process, some planetesimals are left behind and can persist for billions of years after the formation of a planetary system. This is believed to be the principal source of asteroid and Kuiper belt objects, among many other small body populations in the present-day solar system. In addition to providing clues about the process of planet formation, planetesimals can also transport planets themselves, through weak but numerous gravitational interactions. Planetesimals around other stars sometimes collide and generate dust, producing a distinct observational marker that can be used to infer their presence.

In this thesis, I use state of the art N-body simulations to understand the dynamics that govern planetesimal interactions and growth. For the first time, I follow this growth process using bodies with masses comparable to those predicted by planetesimal formation models. Upon doing so, I show that certain dynamical mechanisms involving mean-motion resonances only operate with sufficiently high resolution, and the effects of these mechanisms place a number of constraints on the planet formation process, including tracing the initial sizes of planetesimal formation and using collisionally-generated dust to infer the orbital properties of unseen planets in nearby disks.

I also use these resolution capabilities to directly follow the growth of a system of terrestrial planets, starting from planetesimals. In doing so, I assess the viability of an in-situ formation model for systems of tightly-packed inner planets (STIPs), which appear to be a common outcome of planet formation. I also use these simulation results to train a neural network to generate a larger set of post-planetesimal accretion phase initial conditions, which I leverage to construct a statistical sample of simulated planetary systems for comparison with observations.
}
 
%
% ----- contents & etc.
%
\tableofcontents
\listoffigures
\listoftables 

 
%
% ----- acknowledgments
%
\acknowledgments{% \vskip2pc
  % {\narrower\noindent
  The work presented in this thesis represents the confluence of an incredible amount of care, support and motivation from
  an incredible community of people. Truly, none of this could have happened without them.
  
  First, I'd like to thank my parents for their endless love and support, for helping to show me what I'm truly capable of and for teaching me to prioritize the things that I find personally fulfilling.
  
  I also want to thank my advisor Tom Quinn for being a wonderful mentor and friend over the years. Tom was always patient and kind, but knew exactly how and when to push me to grow as a scientist. I'm also grateful for the many discussions I've had with James Davenport, Aaron Boley and Rory Barnes as both collaborators and mentors.
  
  One thing I never would have expected is that my time in grad school afforded me the opportunity to grow and develop as a musician. I want to thank Brett Morris, Nicole Sanchez and Trevor Dorn-Wallenstein for the incredible experience that was the all-UW-astronomy-grad-student band Night Lunch. Your friendship and the experiences we've shared together are truly the highlight of my time spent living in Seattle.
  
  Finally, I want to thank my partner Justine for her unconditional and limitless love, and for helping me along on the path to figuring out who I truly am. I also want to thank our dogs Tres and Cowboy for sitting on me during the many hours I spent writing this thesis.
  % \par}
}

%
% ----- dedication
%
\dedication{\begin{center}for Grandma Georgi\end{center}}

%
% end of the preliminary pages
 
 
 
%
% ==========      Text pages
%

\textpages

\include{ch1}
\include{ch2}
\include{ch3}
\include{ch4}
\include{ch5}
\include{ch6}

%
% ==========   Bibliography
%
\nocite{*}   % include everything in the uwthesis.bib file
\bibliographystyle{plain}
\bibliography{uwthesis}


\end{document}
